Density-dependent selection and the population ecology of relative fitness

### 2017-11-29 manuscript.tex
l

This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models. Some of the most widely-used models of density-regulated population growth are indeed not compatible: 1) In the logistic model, the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH). It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments. It is not even clear how relative fitness models connect to general ecological models of the form Eq.~\eqref{eq:macgeneral}, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; the total number of territories gained is thus $\Delta_+ N=U(1-e^{-L})$.

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} and its components closely approximate individual-based simulations of the density-dependent lottery model over a wide range of propagule densities $l_i$. Two types are present, one of which is at low frequency. The growth of the low-frequency type relies crucially on the low-density competition term $R_i c_i/\overline{c}$. On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency type, which depends instead on high density territorial victories. Fig.~\ref{fig:simcomp} also shows the breakdown of the classic lottery model at low propagule densities.

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of mean field approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

We now compare MacArthur's claims about selection in crowded environments with our density-dependent lottery model.

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; selection on these saturation densities is commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density. MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} (a) Lotka-Volt}
\end{figure}

Fig.~\ref{fig:LVvslottery} shows two examples of selection when both of two types have the same saturation density $N=K_{11}=K_{22}$. The first shows exlusion by a superior Lotka-Volterra competitor. Note that the trajectories do not follow a line of constant  $N$. The reason for this is that inter-type competition must be stronger than intra-type competition for exclusion to occur, which causes a dip in $N$ over a selective sweep (except for the special case  .

$c$-selection in the density-dependent lottery model, Eq.~\eqref{eq:master}.

\subsection*{Coexistence in constant and cyclical environments}\label{sec:invas}

In the previous section we only considered how $b$, $c$ and $d$ should respond to selection in Grime's environmental extremes, based on invasion fitness. Here we further explore the low frequency behavior of Eq. \eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq. \eqref{eq:master} in a cyclical environment.

In a constant environment, stable coexistence is possible in our extended lottery model. A $b$-specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency-dependent, and so is the importance of competitive ability (Appendix D). This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}; the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model. Coexistence is not possible between $b$- and $d$-specialists in a constant environment (Appendix D).

Now suppose that birth and death rates vary periodically with amplitude sufficent to cause large changes in population density. This example is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months. Along with this seasonal population density cycle, hundreds of polymorphisms exhibit frequency cycles that are in phase with the seasons \citep{bergland_14}. Some of these polymorphisms may be adaptive and potentially millions of years old, suggesting stable coexistence \citep{bergland_14,messer_2016}. Selection on allele frequencies thus occurs on the same time scale as population demography, a situation vastly more complicated than classical sweeps in demographically stable populations \citep{messer_2016}.

The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence. Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different environments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}. Thus, stable coexistence still requires frequency-dependent selection or heterozygote advantage (as is required in a constant environment).

This classical argument overlooks two general mechanisms that promote coexistence in fluctuating environments \citep{messer_2016}. The first is the classic version of the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}). The second is the bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.

Fig.~\ref{fig:fluctuatingselection}a-c shows the behavior of Eq. \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth). Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating. Two types are present, a $b$-specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$-specialist which is better at surviving the winter (lower $d$). Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).

The classic lottery model (Eq. BLAH) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig. \ref{fig:fluctuatingselection}d-f). As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq. \eqref{eq:master} (keeping the other parameters the same; Fig.~\ref{fig:fluctuatingselection}g-i). Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.

\begin{figure}
\centering
\includegraphics[scale=0.7]{fluctuatingselection.pdf}
\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment requires a much greater $b$ advantage in the classic lottery model compared to our density-dependent extension of it when population density is seasonally cyclical. (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer). The $b$-specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$-specialist ($b=0.217$, $d=0.1$) (blue). (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$-specialist does so at a lower rate. Total height (blue+black) is population density $N/T$. (c) Summer favors the $b$ specialist, winter the $d$-specialist, and they stably coexist. (d-f) Same as (a-c) for the classic lottery model; the types no longer coexist. (g-i) Same as (d-f) where now $b = 0.0421$ for the $d$ specialist and the types coexist. For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.} 
\end{figure}

@@

It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}. Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}. Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate at low density and the high density at which saturation occurs (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

%This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection. Yet it is not surprising that $b$ can matter at high densities. In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable. This is a realistic feature of the model. Even if one type is guaranteed to win a territory in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance. For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage. First arrivals are more likely to occur for types with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models. The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing. The logistic model for which $r$ and $K$ are named, does not capture this intuition.

%In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all. The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection. But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines. For instance, contests for territory and mates are predominantly relative. MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption. The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.

@@

There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with a tradeoff between $r$ and the competitive aspect of $K$. This could be driven by a tradeoff between individual size and reproductive rate. To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one type on another type's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each type to produce contestants i.e. $b$ in our model. Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied per iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$). The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.

$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given $r$, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring types which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among types via the propagule density per territory $L$. The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different types. We leave these complications for future work.

Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$-specialists), may also be strongly $d$-selected, which while unintuitive, is consistent with our findings. An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait. Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature. Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.

@@

\section*{Appendix C: Mutant invasion and coexistence in a constant environment}

Here we evaluate the initial growth or decline of mutants in a population with a single resident type, which is in equilibrium. To determine whether coexistence is possible, we check for ``mutual invasion'', that is, we check that type $j$ will invade an $i$-dominated population, but type $i$ will also invade a $j$-dominated population.

Solving for equilibrium when $i$ is the resident ($\Delta n_i = 0$), we have $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} becomes
\begin{equation}
b_i(1-e^{-L})/L-d_i=0.\label{eq:equil}
\end{equation}
This implies $L\approx b_i/d_i$ if $b_i/d_i\gg 1$ and $L\ll 1$ if $b_i/d_i\approx 1$.

Now suppose that a novel mutant $j$, which is initially rare, appears in the population. Then $A_j/R_j\ll 0$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq. \eqref{eq:master}, the mutant lineage's fitness is
\begin{equation}
\Delta n_j/n_j \approx b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
\end{equation}
where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\right)$.

We consider the case of coexistence between a $b$-specialist $i$ and a $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$ and $d_i=d_j$). Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant. Then, when $j$ is dominant, we have $\Delta n_i/n_i=b_i-d_i=b_i-d_j=b_i-b_j>0$. When $i$ is dominant, Eq.  \eqref{eq:highLinvad} applies, where Eq. \eqref{eq:equil} implies $d_j=d_i=b_i(1-e^{-L})/L\approx b_i/L$, and so
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j}{L}\frac{c_j}{c_i}-\frac{b_i}{L}.
\end{equation}
Therefore, coexistence occurs if $c_j/c_i$ is sufficiently large. The analogous argument for $d$- and $c$-specialists ($d_i<d_j$ with $L\gg 1$ when $i$ dominates, $L\ll 1$ when $j$ dominates, and $b_i=b_j$) gives $\Delta n_j/n_j \approx d_i\frac{c_j}{c_i}-d_j$, which again implies coexistence if $c_j/c_i$ is sufficiently large.

For $b$-and $d$-specialists ($c_i=c_j$), we have $\Delta n_j/n_j\approx b_j d_i/b_i-d_j$ when $i$ dominates and $\Delta n_i/n_i\approx b_i d_j/b_j-d_i$ when $j$ dominates. Thus, either $i$ or $j$ grows when rare, but not both, and stable coexistence is not possible in a constant environment.

>>>>

The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In particular, the $f_1$ and $f_2$ nullclines are not even defined. It is clear, however, that constant-$N$ relative fitness models are not compatible with some of the most widely-used models of density-regulated population growth: 1) In the logistic model, the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys 
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} and its components closely approximate simulations of the density-dependent lottery model over a wide range of propagule densities. Two types are present, one of which is at low frequency. The growth of the low-frequency type relies crucially on the low-density competition term $R_i c_i/\overline{c}$. On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency type, which depends instead on high density territorial victories. Fig.~\ref{fig:simcomp} also shows the breakdown of the classic lottery model at low propagule densities.

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of mean field approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

We now compare the density-dependent lottery model from the previous section with MacArthur's claims about selection in crowded environments.

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
\frac{d n_1}{dt} = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1 = f_1(n_1,n_2)\nonumber\\
\frac{d n_2}{dt} = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2 = f_2(n_1,n_2)\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density. MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Selection between types with identical saturation density but different inter-type competitive ability. (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$. Trajectories do not follow the line $N=K_{11}=K_{22}$. (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=0.1$ and $c_1/c_2=5$. Trajectories converge on the line $N=K_{11}=K_{22}$.}
\end{figure}

Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$. We thus now restrict our attention to the case $K_{11}=K_{22}$. Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density. Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$). Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$. This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density. Selection then occurs purely along this line, uncoupled from the density regulation of $N$. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all. In other words, $c$ is a zero-sum relative fitness trait embedded within a non-zero-sum fitness model.

Thus, we can conclude that relative fitness models are perfectly consistent with MacArthur's claim that ``fitness is $K$'' in crowded populations, bearing in mind his broad use of the symbol ``$K$''. However, inconsistency remains with standard growth models such as the Lotka-Volterra model. We revisit this issue in the Discussion.

%, we will evaluate the broader ecological plausibility of relative fitness models from the basis of our density-dependent lottery model.

@@

If s is small it doesn't matter anyway

%The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection. That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types. Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection. But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines. For instance, contests for territory and mates are predominantly relative. MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption. The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.

@@

%To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one type on another type's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each type to produce contestants i.e. $b$ in our model.

@@

\section*{Appendix C: Total density in the Lotka-Volterra competition model}

Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density.

We assume $\alpha_{11}=\alpha_{22}$ and $N=1/\alpha_{11}$ and check whether it is then possible for $\frac{dN}{dt}$ to be zero in the sweep ($n_1,n_2 \neq 0$). Substituting these conditions into Eq.~\eqref{eq:LV}, we obtain 
\begin{align}
\frac{d n_1}{dt} = -r_1(\alpha_{12}-\alpha_{11})n_1n_2 \nonumber\\
\frac{d n_2}{dt} = -r_2(\alpha_{21}-\alpha_{22})n_1n_2
\end{align}
Adding these together, $\frac{dN}{dt}$ can only be zero if 
\begin{equation}
r_1(\alpha_{12}-\alpha_{11})+r_2(\alpha_{21}-\alpha_{22})=0. \label{eq:constNcondition}
\end{equation}
To get some intuition for Eq.~\eqref{eq:constNcondition}, suppose that a mutant arises with improved competitive ability but identical intrinsic growth rate and saturation density ($r_1=r_2$ and $\alpha_{11}=\alpha_{22}$). This could represent a mutation to an interference competition trait, for example \citep{gill_1974}. Then, according the above condition, for $N$ to remain constant over the sweep, the mutant must find the wildtype more tolerable than itself by exactly the same amount that the wildtype finds the mutant less tolerable than itself. This condition, and Eq.~\eqref{eq:constNcondition} more generally, are so restrictive that we can conclude that selective sweeps in the Lotka-Volterra competition model will generally involve non-constant $N$.
